Abstractions for Multi-Sorted Substitutions

### 2023-08-26 paper/paper.tex
Update.

\newcommand*\ACode[1]{\AgdaFontStyle{#1}}
\newcommand*\AField[1]{\AgdaField{#1}}
\newcommand*\ACon[1]{\AgdaInductiveConstructor{#1}}
\newcommand*\AKw[1]{\AgdaKeyword{#1}}
\newcommand*\ADef[1]{\AgdaFunction{#1}}

@@

  The term type \ACode{S \ACon{⊢} s} is then indexed by its sort
  \ACode{s} and the sorts of its free variables \ACode{S}.
  For example, \ACon{[𝕥, 𝕥] ⊢ 𝕖} is the type of expressions (\ACon{𝕖}) with two free type-variables (\ACon{𝕥}).
  \newpage
  \AppCode\FSyntax
  The notation \ACon{\_⊢\_} is often used for terms in
  intrinsically-typed languages. This is no accident: in effect, we
  defined an intrinsically-typed language with the twist that the
  typing relation assures exactly that the syntactic categories are followed.
  Sorts \ACode{s} correspond to types, and lists of sorts \ACode{S}
  correspond to type environments.

@@

  Contrary to the usual intrinsic typing, we limit the environment
  \ACode{S} to sorts with index \ACon{Var}.
  This ensures that it is impossible to construct kind variables, as
  it is not possible to extend \ACode{S} with the sort of kinds
  \ACon{𝕜}.

  \subsection{A Structure for Multi-Sorted Syntax}
  The multi-sorted syntax makes it easy to capture general syntax definitions
  in a structure:
  \LibCode\KTerms
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \AField{`\_} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FTerms

@@

  \subsection{The Need for Renamings}
  A problem with mechanizing languages with DeBruijn-indices is that
  the operation of applying a substitution \ACode{σ} to a term
  \ACode{t} cannot be defined directly by structural recursion on
  \ACode{t}.
  If \ACode{t} is a term like \ACode{\ACon{λx} e},
  where the subterm \ACode{e} is under a binder, then the variables
  in \ACode{e} are shifted, i.e.\ the variable \ACon{zero} refers
  to the variable from the binder, whereas a variable \ACode{\ACon{suc} x}
  refers to the variable \ACode{x} from outside of the binder.
  As the terms contained in \ACode{σ} may themselves contain free variables,
  they need to be shifted, too, before they can be substituted
  for variables in \ACode{e}. 
  Shifting the variables in a term is itself a substitution, which maps
  each variable \ACode{x} to the term \ACode{\ACon{`} (\ACon{suc} x)}.
  However, applying this shifting substitution to the terms in
  \ACode{σ} is not structurally recursive, as the terms in \ACode{σ}
  are not subterms of \ACode{\ACon{λx} e}.

@@

  As we intend to have exactly two kit instances, we choose names of
  the form \ACode{x/y}, where \ACode{x} is the name we choose for the
  variable instance, and \ACode{y} is the name we choose for the term
  instance.
  For example the parameter type \ACode{\_∋/⊢\_} will be instanciated
  to \ACon{\_∋\_} for the variable kit, and to \AField{\_⊢\_} for the
  term kit.

  A kit provides the following operations:
  \begin{itemize}
  \item 
    \AField{id/`} converts a variable \ACode{S \ACon{∋} s} into a
    \ACode{S \AField{∋/⊢} s}. For the variable kit, \ACode{S ∋/⊢ s} is instanciated to \ACode{S \ACon{∋} s}, so this
    operation is the identity. For the term kit, \ACode{S ∋/⊢ s} is
    instanciated to \ACode{S \AField{⊢} s}, so this operation is the
    variable constructor \AField{`\_}.
  \item 
    \AField{`/id} converts a \ACode{S ∋/⊢ s} into a term \ACode{S \AField{⊢} s}
    and is dual to the \AField{id/`} operation.
  \item 
    \AField{wk} shifts the DeBruin-Indices in a \ACode{S ∋/⊢ s}.
    The new, unused variable \ACon{zero} can assume any sort \ACode{s'}.
    For variables \AField{wk} is the successor \ACon{suc}.
    For terms \AField{wk} means applying a shifting renaming to the term.
  \end{itemize}

  The operations need to satisfy the following properties:
  \begin{itemize}
  \item \AField{`/`-is-`} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \AField{`\_}.
  \item \AField{`/id-injective} and \AField{id/`-injective} state that
    \AField{`/id} and \AField{id/`} are injective. This property
    follows easily, as both the identity function and the variable
    constructor are injective.
  \item \AField{wk-id/`} characterizes the behaviour of the \AField{wk} function.
    Injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or a terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map}.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \AField{wk} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACon{zero} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \AField{wk} on terms:
    \LibCode\KWeaken
  \end{itemize}

  We also specify additional notation, which makes it less cumbersome
  to explicitly pass a particular Kit to those definitions:
  \LibCode\KKitNotation

  The only task left now is to create the two instances for the kits
  and define how a map can be applied to a term.
  While the variable kit can be defined as expected, the term kit
  raises a dependency problem: The \AField{wk} operation for terms requires us
  to already know how to apply a renaming to term.
  To solve this problem, McBride came up with a clever solution:
  Applying a map to a term, can be defined without even having any Kit-instances yet.
  Afterwards we can first define the variable kit, and in the
  definition of the term kit, we are then already allowed to apply renamings.
  After the definition of the term-kit we are then allowed to apply
  both renamings and substitutions as both kits defined.

  \LibCode\KKitVar

  \subsection{Library}
  \LibCode\KKitOpenInst
  \LibCode\KEq
  \LibCode\KFunExt
  \LibCode\KIdLift
  % \LibCode\KIdLiftProof
  \LibCode\KTraversal
  \LibCode\KKitVar
  \LibCode\KKitTerm
  \LibCode\KKitOpen
  \subsection{System F}
  \AppCode\FTraversalOp
  \AppCode\FTraversalId
  \AppCode\FTraversalIdProofInteresting
  % \LibCode\FTraversalIdProofRest
  \AppCode\FTraversal

  \section{Map Composition}
  \subsection{Library}
  \LibCode\KWkKit
  \LibCode\KWkKitInstances
  \LibCode\KComposeKit
  \LibCode\KComposition
  \LibCode\KComposeKitAp
  % \LibCode\KComposeKitApProof
  \LibCode\KDistLiftCompose
  % \LibCode\KDistLiftComposeProof
  \LibCode\KComposeKitNotation
  \LibCode\KComposeTraversal
  \LibCode\KCommLiftWeaken
  % \LibCode\KCommLiftWeakenProof
  \LibCode\KCommLiftWeakenTraverse
  % \LibCode\KCommLiftWeakenTraverseProof
  \LibCode\KComposeKitInstances
  \LibCode\KComposeKitInstancesConcrete
  \LibCode\KWeakenCancelsSingle
  % \LibCode\KWeakenCancelsSingleProof
  \LibCode\KWeakenCancelsSingleTraverse
  % \LibCode\KWeakenCancelsSingleTraverseProof
  \LibCode\KDistLiftSingle
  % \LibCode\KDistLiftSingleProof
  \LibCode\KDistLiftSingleTraverse
  % \LibCode\KDistLiftSingleTraverseProof
  \subsection{System F}
  \AppCode\FAssoc
  \AppCode\FAssocProofInteresting
  % \AppCode\FAssocProofRest
  \AppCode\FComposeTraversal

>>>>

% \setlength\fboxsep{0pt}
% \usepackage{realboxes}
% \definecolor{mygray}{rgb}{0.95,0.95,0.95}
% \newcommand*\ACode[1]{\Colorbox{mygray}{\AgdaFontStyle{#1}}}

% \newcommand*\ACode[1]{\AgdaFontStyle{#1}}

% \newcommand*\ACode[1]{\texttt{#1}}

\definecolor{mygray}{rgb}{0.4,0.4,0.4}
\newcommand*\ACode[1]{\AgdaFontStyle{\textcolor{mygray}{#1}}}

\newcommand*\AField[1]{\AgdaField{#1}}
\newcommand*\ACon[1]{\AgdaInductiveConstructor{#1}}
\newcommand*\AKw[1]{\AgdaKeyword{#1}}
\newcommand*\ADef[1]{\AgdaFunction{#1}}

@@

  The term type \ACode{S \ACon{⊢} s} is then indexed by its sort
  \ACode{s} and the sorts of its free variables \ACode{S}.
  For example, \ACode{\ACon{[𝕥, 𝕥] ⊢ 𝕖}} is the type of expressions
  (\ACode{\ACon{𝕖}}) with two free type-variables (\ACode{\ACon{𝕥}}).
  \newpage
  \AppCode\FSyntax
  The notation \ACode{\ACon{\_⊢\_}} is often used for terms in
  intrinsically-typed languages. This is no accident: in effect, we
  defined an intrinsically-typed language with the twist that the
  typing relation assures exactly that the syntactic categories are followed.
  Sorts \ACode{s} correspond to types, and lists of sorts \ACode{S}
  correspond to type environments.

@@

  Contrary to the usual intrinsic typing, we limit the environment
  \ACode{S} to sorts with index \ACode{\ACon{Var}}.
  This ensures that it is impossible to construct kind variables, as
  it is not possible to extend \ACode{S} with the sort of kinds
  \ACode{\ACon{𝕜}}.

  \subsection{A Structure for Multi-Sorted Syntax}
  The multi-sorted syntax makes it easy to capture general syntax definitions
  in a structure:
  \LibCode\KTerms
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \ACode{\AField{`\_}} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FTerms

@@

  \subsection{The Need for Renamings}
  A problem with mechanizing languages with DeBruijn-indices is that
  the operation of applying a substitution \ACode{σ} to a term
  \ACode{t} cannot be defined directly by structural recursion on
  \ACode{t}.
  If \ACode{t} is a term like \ACode{\ACon{λx} e},
  where the subterm \ACode{e} is under a binder, then the variables
  in \ACode{e} are shifted, i.e.\ the variable \ACode{\ACon{zero}} refers
  to the variable from the binder, whereas a variable \ACode{\ACon{suc} x}
  refers to the variable \ACode{x} from outside of the binder.
  As the terms contained in \ACode{σ} may themselves contain free variables,
  they need to be shifted, too, before they can be substituted
  for variables in \ACode{e}. 
  Shifting the variables in a term is itself a substitution, which maps
  each variable \ACode{x} to the term \ACode{\ACon{`} (\ACon{suc} x)}.
  However, applying this shifting substitution to the terms in
  \ACode{σ} is not structurally recursive, as the terms in \ACode{σ}
  are not subterms of \ACode{\ACon{λx} e}.

@@

  As we intend to have exactly two kit instances, we choose names of
  the form \ACode{x/y}, where \ACode{x} is the name we choose for the
  variable instance, and \ACode{y} is the name we choose for the term
  instance.
  For example the parameter type \ACode{\_∋/⊢\_} will be instanciated
  to \ACode{\ACon{\_∋\_}} for the variable kit, and to \ACode{\AField{\_⊢\_}} for the
  term kit.

  A kit provides the following operations:
  \begin{itemize}
  \item 
    \ACode{\AField{id/`}} converts a variable \ACode{S \ACon{∋} s} into a
    \ACode{S \AField{∋/⊢} s}. For the variable kit, \ACode{S ∋/⊢ s} is instanciated to \ACode{S \ACon{∋} s}, so this
    operation is the identity. For the term kit, \ACode{S ∋/⊢ s} is
    instanciated to \ACode{S \AField{⊢} s}, so this operation is the
    variable constructor \ACode{\AField{`\_}}.
  \item 
    \ACode{\AField{`/id}} converts a \ACode{S ∋/⊢ s} into a term \ACode{S \AField{⊢} s}
    and is dual to the \ACode{\AField{id/`}} operation.
  \item 
    \ACode{\AField{wk}} shifts the DeBruin-Indices in a \ACode{S ∋/⊢ s}.
    The new, unused variable \ACode{\ACon{zero}} can assume any sort \ACode{s'}.
    For variables \ACode{\AField{wk}} is the successor \ACode{\ACon{suc}}.
    For terms \ACode{\AField{wk}} means applying a shifting renaming to the term.
  \end{itemize}

  The operations need to satisfy the following properties:
  \begin{itemize}
  \item \ACode{\AField{`/`-is-`}} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \ACode{\AField{`\_}}.
  \item \ACode{\AField{`/id-injective}} and \ACode{\AField{id/`-injective}} state that
    \ACode{\AField{`/id}} and \ACode{\AField{id/`}} are injective. This property
    follows easily, as both the identity function and the variable
    constructor are injective.
  \item \ACode{\AField{wk-id/`}} characterizes the behaviour of the \ACode{\AField{wk}} function.
    Injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map}.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \ACode{\AField{wk}} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACode{\ACon{zero}} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \ACode{\AField{wk}} on terms:
    \LibCode\KWeaken
  \item
    Maps have an extensional equality:
    \LibCode\KEq
  \item
    For simplicity we postulate functional extensionality:\footnote{
      The actual implementation does \emph{not} use any postulates.
      Appendix TODO explains the workaround.
    }
    \LibCode\KFunExt
  \item
    Lifting the identity map is a (larger) identity map:
    \LibCode\KIdLift
    % \LibCode\KIdLiftProof
  \end{itemize}

  We also specify notation, which makes it less cumbersome
  to explicitly pass a particular Kit to those definitions:
  \begin{itemize}
  \item \ACode{S \ADef{∋/⊢[} K \ADef{]} s} is \ACode{S ∋/⊢ s} for some \ACode{\ACon{Kit}} \ACode{K}; and
  \item \ACode{S \ADef{–[} K \ADef{]→} s} is \ACode{S \ADef{→ₖ} s} for some \ACode{\ACon{Kit}} \ACode{K}.
  \end{itemize}

  The operation of applying a map to a term depends on the concrete
  structure of a term, so we define a structure for it:
  \LibCode\KTraversal
  The fields of this structure are:
  \begin{itemize}
  \item
    \ACode{t \AField{⋯} ϕ} applies the map \ACode{ϕ} (a renaming or
    substitution) to the term \ACode{t}.
  \item
    \ACode{\AField{⋯-var}} states that applying a map \ACode{ϕ} to a variable
    term \ACode{\AField{`} x}, is the same as applying \ACode{ϕ} to
    the variable \ACode{x}, and then converting the result from \ACode{S₂ ∋/⊢ S}
    to \ACode{S₂ ⊢ S} via \ACode{\AField{id/`}}.
  \item
    \ACode{\AField{⋯-id}} states that applying the identity map
    \ACode{id} to a term does not change the term.
  \end{itemize}

  Now we are finally ready to define the actual kit instances.
  The variable kit definition is straightforward:
  \LibCode\KKitVar

  The term kit requires both the variable kit and the
  \ACode{\ACon{Traversal}} to be defined, because shifting a term \ACode{t} with \ACode{\AField{wk}}
  means applying the shifting renaming to the term.
  Hence, we define the term kit in the record module of \ACode{\ACon{Traversal}}:
  \LibCode\KKitTerm

  % The only task left now is to create the two instances for the kits
  % and define how a map can be applied to a term.
  % While the variable kit can be defined as expected, the term kit
  % raises a dependency problem: The \ACode{\AField{wk}} operation for terms requires us
  % to already know how to apply a renaming to term.
  % To solve this problem, McBride came up with a clever solution:
  % Applying a map to a term, can be defined without even having any Kit-instances yet.
  % Afterwards we can first define the variable kit, and in the
  % definition of the term kit, we are then already allowed to apply renamings.
  % After the definition of the term-kit we are then allowed to apply
  % both renamings and substitutions as both kits defined.

  Note that all definitions so far are library code. To make them
  available for our System F example, all we need to do is
  to create an instance of \ACode{\ACon{Traversal}}.

  We start with the operation of applying a map to a term:
  \AppCode\FTraversalOp
  The interesting cases are those with variables and binders:
  \begin{itemize}
  \item 
    In the variable case \ACode{(\ACon{`} x) \ADef{⋯} ϕ}, we first apply the map
    \ACode{ϕ} to the variable \ACode{x}. If \ACode{ϕ} is a renaming,
    we get back a variable and need to apply the variable constructor
    \ACode{\ACon{`\_}} to it. If \ACode{ϕ} is a substitution,
    we get back a term that we can use directly.
    Hence, the use of \ACode{`/id}.
  \item 
    In cases where the operation needs to go under a binder, like
    \ACode{(\ACon{λx} e) \ADef{⋯} ϕ}, we lift the map using \AField{\_↑\_} before
    we apply it to the subterm, to account for the bound variable.
  \end{itemize}

  \AppCode\FTraversalId
  \AppCode\FTraversalIdProofInteresting
  % \LibCode\FTraversalIdProofRest
  \AppCode\FTraversal

  \ExampleCode\FExampleTrav

  % \subsection{Unused}
  % \LibCode\KKitOpenInst
  % \LibCode\KKitOpen

  \section{Map Composition}
  An operation we did not cover in the previous section is the
  composition of substitutions.
  This operation and its properties are relevant in numerous places, e.g.\
  TODO, which are also relevant for System F.

  As we defined substitution in terms of renamings, we now have to
  consider all four compositions between renamings and substitutions,
  and their properties. The problem is again that they are dependent
  on each other, i.e.\ composition between two substitutions relies on
  the compositions between renamings and substitutions, which rely on
  compositions between two renamings.

  McBride et al.\cite{DBLP:journals/jar/BentonHKM12} addressed this
  issue by duplicating the definitions and using tactics to reduce
  boilerplate in proofs.

  In contrast we define a structure similar to Kits, which allows us
  to abstract over all four compositions and use the same trick as
  before to eliminate the dependencies.
  This has the benefit that a user of our library has to write less
  boilerplate and also that we gain more expressiveness, which helps
  us later to unify type preservation of all substitution and renaming
  operations into a single lemma.

  We first define a \ACode{\ACon{WkKit}}, which captures properties
  between maps and map application, which we could not include in the
  \ACode{\ACon{Kit}} definition, as we already require
  \ACode{\ACon{Kit}}s to define a Traversal. We then create exactly
  two instances of the \ACode{\ACon{WkKit}} for the variable and term
  kit, respectively. Both the structure and the instances are part of
  the library and do not require application code.

  Using the \ACode{\ACon{WkKit}}, we then define
  the \ACode{\ACon{ComposeKit}} and \ACode{\ACon{ComposeTraversal}}
  structures, which serve analogous purposes to \ACode{\ACon{Kit}}
  and \ACode{\ACon{Traversal}} from before:
  the library provides a \ACode{\ACon{ComposeKit}} for each of the
  four compositions, whereas the \ACode{\ACon{ComposeTraversal}}
  structure needs to be instanciated by the application with
  lemmas parameterized over arbitrary \ACode{\ACon{ComposeKit}}s.
  This gives the user a composition operation
  \ACode{\AField{\_·ₖ\_}} between arbitrary maps and corresponding
  lemmas, instead of multiple definitions and lemmas about concrete
  maps, enabling us to reason on a more abstract level and to avoid
  further boilerplate.

  \subsection{WkKits}

  \LibCode\KWkKit
  \LibCode\KWkKitInstances

  \subsection{Composition}

  \LibCode\KComposeKit
  \LibCode\KComposition
  \LibCode\KComposeKitAp
  % \LibCode\KComposeKitApProof
  \LibCode\KDistLiftCompose
  % \LibCode\KDistLiftComposeProof
  \LibCode\KComposeKitNotation
  \LibCode\KComposeTraversal
  \LibCode\KCommLiftWeaken
  % \LibCode\KCommLiftWeakenProof
  \LibCode\KCommLiftWeakenTraverse
  % \LibCode\KCommLiftWeakenTraverseProof
  \LibCode\KComposeKitInstances
  \LibCode\KComposeKitInstancesConcrete
  \LibCode\KWeakenCancelsSingle
  % \LibCode\KWeakenCancelsSingleProof
  \LibCode\KWeakenCancelsSingleTraverse
  % \LibCode\KWeakenCancelsSingleTraverseProof
  \LibCode\KDistLiftSingle
  % \LibCode\KDistLiftSingleProof
  \LibCode\KDistLiftSingleTraverse
  % \LibCode\KDistLiftSingleTraverseProof

  \subsection{System F}
  \AppCode\FAssoc
  \AppCode\FAssocProofInteresting
  % \AppCode\FAssocProofRest
  \AppCode\FComposeTraversal
